\documentclass{article}
\usepackage{html}
\usepackage{hyperref}
\usepackage{tabularx}
\usepackage[top=0.75in,bottom=0.75in,left=1.25in,right=1.25in]{geometry}
\usepackage{array}
\usepackage{graphics}
\hypersetup{
   pdfinfo={ %
      Title={Z4800 Documentation} %
   }, %
   bookmarks=true,         % show bookmarks bar?
   unicode=false,          % non-Latin characters in Acrobat’s bookmarks
   pdftoolbar=true,        % show Acrobat’s toolbar?
   pdfmenubar=true,        % show Acrobat’s menu?
   pdffitwindow=true,      % window fit to page when opened
   pdfnewwindow=true,      % links in new window
   colorlinks=true,        % false: boxed links; true: colored links
   linkcolor=blue,         % color of internal links
   citecolor=green,        % color of links to bibliography
   filecolor=blue,         % color of file links
   urlcolor=blue           % color of external links
}

\newcommand{\intref}[2]{%
   \latexhtml{%
      \hyperlink{#1}{#2 (\S\ref*{#1}, p.\pageref*{#1})}%
   }%
   {%
      \htmlref{#2}{#1}%
   }%
}

\newcommand{\quotes}[1]{\latexhtml{``#1''}{"#1"}}
\newcommand{\tm}{\latexhtml{\texttrademark}{\begin{rawhtml}&\#8482;\end{rawhtml}}}
\newcommand{\rtm}{\latexhtml{\textregistered}{\begin{rawhtml}&reg;\end{rawhtml}}}

\title{Z4800 Platform Documentation}
\author{Will Simoneau}

\begin{document}
\maketitle

\begin{rawhtml}
<div id="author_hack"></div>
\end{rawhtml}

The Z4800 platform is an open-source (mostly GPLv2) project consisting of a
synthesizable soft-core CPU plus supporting hardware and software. The included
reference designs allow construction of a complete FPGA computer complemented
by a full-featured operating system capable of running many real-world
programs. This platform can be used for many purposes, particularly benchmark
studies; the processor core can be modified and has an extendable performance
counter system.

The Z4800 CPU itself is a clean-room design capable of executing some binaries
compiled for a variant of the MIPS\rtm{} R4000\tm{} processor. The main
differences are that it does not implement the 64-bit or floating-point
instructions, and that its native endianness is little-endian.  Aside from
these differences, most existing user-mode and kernel-mode software will work.

A fully functional Linux kernel is supported; source code for the modified
kernel is included. These modifications are mostly to support the peripherals
on the FPGA boards -- the Z4800 CPU functions enough like an R4K that the
existing Linux/mips R4K code can be used directly. SMP (multiprocessor) support
is included.

To download the complete Z4800 source code, run the following command. You will
need at least Git 1.5.3 because this project uses submodules.
\begin{verbatim}
git clone --recursive git://trogdor.ele.uri.edu/z4800.git
\end{verbatim}

A paper about this project,
\htmladdnormallink{%
      \emph{%
         An FPGA-based Multi-Core Platform for Testing and Analysis of
         Architectural Techniques%
      }%
   }%
   {http://www.ele.uri.edu/~simoneau/ispass2012.pdf}%
, was written for and accepted to the
\emph{%
   \htmladdnormallink{%
      2012 IEEE International Symposium on Performance Analysis of Systems
      and Software%
   }%
   {http://ispass.org/ispass2012}%
}.

\smallskip

\subsection{To-Do List}
In no particular order:
\begin{itemize}
\item Publish public documentation on the Z4800 CPU internals
\item Add screenshots to the documentation to show example output
\item Create and publish virtual machine images providing a canned development environment
\item Create and publish tarball providing a canned root-FS for target board (DE2-115, etc.)
\item Publish public version of benchmarking infrastructure, with accompanying documentation
\end{itemize}

\subsection{Legal Notice}
{
   \footnotesize
   MIPS and R4000 are trademarks or registered trademarks of MIPS Technologies,
   Inc. in the United States and other countries. The Z4800 project is neither
   licensed by nor endorsed in any way by MIPS Technologies, Inc. We only claim
   to have limited binary compatibility; we claim neither full compatibility
   nor full compliance with any MIPS\rtm{} products and/or specifications.
}

\subsection{Contacts}

The author of the Z4800 source code is Will Simoneau
(\htmladdnormallink{simoneau@ele.uri.edu}{mailto:simoneau@ele.uri.edu}). He may
be contacted for any technical questions.
The ISPASS 2012 paper was co-authored with \htmladdnormallink{Resit
Sendag}{http://www.ele.uri.edu/faculty/sendag/}
(\htmladdnormallink{sendag@ele.uri.edu}{mailto:sendag@ele.uri.edu}).

\input{sec_dev_env}
\input{sec_files}

\end{document}
